Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-28 demystify.tex
updates

\begin{table}
  \centering
  \small
  \caption{Different types of PhD funding. All cover tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Your TA is paid through the school or department, i.e., they hire you to help teach.
As a TA, you spend up to 20 hrs/week and help with classes (e.g., grading or teaching labs/recitation).
During a semester, a TA might work with several courses and professors (not necessarily their adviser).  TA funding is not typically available during the summer, which has few courses.

\paragraph{How to get TA?}  Unless you have other funding such as RA or Fellowships, TA is typically default for CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

\begin{commentbox}[Vu:]
  At GMU CS, PhD students admitted TA have 4 years of GTA guaranteed, and in also receive a stipend for the first summer.
\end{commentbox}

@@

It is important to note that RA is never guaranteed as it depends on the funding situation of the prof. So you should also pay attention to TA, which is a good backup plan (remember, typically TA and RA have similar benefits). This means you should also ask if TA is readily available for PhD students in the program.

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous and gives pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipend (including summer) and open doors for job opportunities such as internships.

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract admitted students. You likely will not need to do anything and the adcom will recommend such fellowships to admitted students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

For external fellowships including those from the US government and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizen and permanent residents, for underrepresented minority (URM) students (\autoref{sec:urm}), and for students from specific countries. In short, there are many fellowships available, so you should look for them and apply.

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website, and you should put it on your CV.

>>>>

\begin{table}
  \centering
  \small
  \caption{Different types of PhD funding. All cover tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Likely                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Your TA is paid through the school or department, i.e., they hire you to help teach.
As a TA, you spend up to 20 hrs/week and help with classes (e.g., grading or teaching labs/recitation).
During a semester, a TA might work with several courses and professors (not necessarily their adviser).  TA funding is \emph{not} typically available during the summer, which has few courses.

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically default for CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

\begin{commentbox}[Vu:]
  At GMU CS, admitted PhD students have 4 years of GTA guaranteed, and also receive a stipend for the first summer.
\end{commentbox}

@@

It is important to note that RA is \emph{never guaranteed} as it depends on the funding situation of the prof. So you should also pay attention to TA, which is a good backup plan (remember, typically TA and RA have quite similar benefits). This means you should also check if TA is readily available for PhD students in the program.

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, giving pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipend (including summer) and open doors for job opportunities such as internships.

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract students. You likely will not need to do anything and the adcom will recommend such fellowships to strong students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

For external fellowships including those from the US government (e.g., NSF, DOD) and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizen and permanent residents, for underrepresented minority (URM) students (\autoref{sec:urm}), and for students from specific countries. There are many fellowships available, so you should look for them and apply.

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, school will still mention you on their website, and you should put it on your CV.
